Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2018-06-07 articles/formalism/sections/real_space_formalism.tex
added definition y_dagger and introduced normalization in all necessary places. Revised some discussion with added clarity. Changes marked in red text left some comments to be addressed

\section{Polarization primer} \label{sec:pol-primer}
%\subsection{Constructing $E/B$ fields without harmonic space}
\subsection{A heuristic argument for real space construction of E/B fields on the sphere}
%\comment{Need a better title for the section that gives the heuristic argument}
%\textit{Alternative for bringing the heuristic argument earlier.}

 CMB polarization is measured in terms of Stokes parameters, time averages of the linear polarization of the electric field along cartesian axes perpendicular to the line of sight.\footnote{Throughout we use the conventions of HEALPix \cite{healpix_primer}, measuring the polarization angle East of South.} Thus Stokes $Q$ and $U$ depend on the choice of the local coordinate system, and a rotation by an angle $\psi$ around the line of sight transforms them as
%
\beq \label{eq:qu-rot}
\fqu' = \begin{bmatrix} \cos{2 \psi} &  \sin{2 \psi} \\ -\sin{2\psi} & \cos{2 \psi} \end{bmatrix} \fqu \,.
\eeq
Equivalently the object $_{\pm 2}\bar{X}(\hat{n}) = Q(\hat{n}) \pm i U (\hat{n})$ transforms as ${}_{\pm 2}f' = e^{\mp 2i\psi} {}_{s}f$ and hence forms a spin-2 field \cite{Zaldarriaga1997}.

@@

We consider the contribution to a scalar field at $\hat n_e$ from the polarization field at $\hat n_q$. \fig{fig:euler_angles} shows that the transformation of the local coordinate system between the two positions can be described by a \revisit{counter-clockwise rotation around the local $\hat n_q$ by angle $\alpha$, parallel transport by angle $\beta$ along the shorter geodesic and a counter-clockwise rotation around $\hat n_q$ by $-\gamma$.  This corresponds to a rotation by Euler angles $(\alpha,\beta,-\gamma)$ in the $z-y_1-z_2$ convention.\footnote{The Euler angles in the more standard $z-y-z$ convention are related to those in the $z-y_1-z_2$ convection by the following rule: $(\alpha,\beta,\gamma)_{z-y-z} =(\gamma,\beta,\alpha)_{z-y_1-z_2}$ \cite{varshalovich}.}}\\
%
\begin{figure}%[!hbt]
\centering
\includegraphics[width=0.5\columnwidth]{euler.pdf}
\caption{This figure depicts the Euler angles in the z-y1-z2 convention. The cartesian coordinates shown in dark green are those that lie in the tangent plane at location $\hat{n}_q = (\theta_q, \phi_q)$ while those shown in dark red are the ones that lie in the tangent plane at location $\hat{n}_e = (\theta_e, \phi_e)$. The blue coordinates at different locations are representative of the parallel transport along the geodesic connection the two locations $\hat{n}_q$ \& $\hat{n}_e$ on the sphere.}
\label{fig:euler_angles}
\end{figure}
%
Rotating the cartesian coordinates in the tangent plane at location $\hat{n}_q$ by an angle $\psi$ about the local $\hat{z}_q$ axis, the Stokes parameters in the new coordinate system relate to those in the original coordinate system as:
$\mathcal{R}_{\hat{z}_q}(\psi)[{}_{+2}X(\hat{n}_q)] =  {}_{+2}X(\hat{n}_q) e^{-i2\psi} $.
This same rotation by $\psi$ alters the Euler angle $\alpha_{qe}$ (that aligns the $x$-axis at $\hat{n}_q$ along the geodesic to the location $\hat{n}_e$): $\mathcal{R}_{\hat{z}_q}(\psi)[\alpha_{qe}] = \alpha_{qe} - \psi$.  Therefore one can see that $\mathcal{R}_{\hat{z}_q}(\psi)[e^{-i2\alpha_{qe}}] =  e^{-i2\alpha_{qe}} e^{i2\psi}$.

Given these transformation properties, the combination ${}_{+2}X(\hat{n}_q)e^{-i2\alpha_{qe}}$ is invariant under rotations and must be spin-0 by definition:
\beq
\mathcal{R}_{\hat{z}_q}(\psi)[{}_{+2}X(\hat{n}_q)e^{-i2\alpha'_{qe}}] = {}_{+2}X(\hat{n}_q) e^{-i2\alpha_{qe}}
\eeq
\comment{I'm having some trouble understanding the next 2 sentences.} The real part of the function is constructed by product of functions $(Q\cos{2\alpha}, U\sin{2 \alpha})$ having the same parity and hence the real part must have even parity.  At the same time, the imaginary part of the function is constructed by multiplying functions $(Q\sin{2 \alpha}, U\cos{2\alpha})$ of opposite parity and hence must have an odd parity. Therefore we can make the association that contributions to  $(E+iB)(\hat n_e)$ must be proportional to $ {}_{+2}X(\hat{n}_q) e^{-i2\alpha_{qe}}$.

The same rotation $\mathcal{R}_{\hat{z}_q}(\psi)$ leaves the Euler angle $|\beta_{qe}|$ unaltered (it measures the angular distance between the points).  Thus we can conclude that the contribution to $(E+iB)(\hat n_e)$ from the position $n_q$ must have the form 
\beq
{}_{+2}X(\hat{n}_q) f(\beta_{qe})  e^{-i2\alpha_{qe}}
\eeq
for some real function $f$.  Note that when the two locations coincide ($\beta_{qe}=0$) then  $\alpha_{qe}=0,2\pi,4\pi,\dots$, implying $E + iB \propto Q+iU$.  This is a contradiction because $Q+iU$ does not transform as a spin-0 field under local rotations, and so we must have $f(\beta_{qe} = 0 ) = 0$.   Hence the $E/B$ fields are necessarily non-local.  A similar contradiction arises when the two locations are diametrically opposite, $\beta_{qe} = \pi$, and so $f(\beta_{qe} = \pi ) = 0$.  Any such function $f$ will let us construct $E/B$-like scalar fields.  Below we derive the particular one that gives rise to our familiar $E/B$ modes.

@@

The standard construction of $E/B$ fields depend on the spin-raising and -lowering operators, and is usually carried out in harmonic space.
%
The spin-raising operator ($\eth$), applied to a field of spin-s $_{s}g$, results in a fields with spin-$(s+1)$: $(\eth _{s}g)' = e^{-i(s+1)\psi}(\eth _{s}g)$  \cite{goldberg67}.  The complementary spin-lowering operator $(\bar{\eth})$  similarly results with $(\bar{\eth} _{s}g)' = e^{-i(s-1)\psi}(\bar{\eth} _{s}g)$.  The complex spin-0 scalar now arise from  the spin-2 fields ${_{\pm 2}X}$ as follows,
%
\begin{subequations}\label{eq:ebdef}
\beqry
\mathcal{E}(\hat{n}) + i \mathcal{B}(\hat{n}) &=& -\bar{\eth}^2 _{+ 2}\bar{X}(\hat{n}) \,,\label{eq:ebdef_lower}\\
\mathcal{E}(\hat{n}) - i \mathcal{B}(\hat{n}) &=& -{\eth}^2 _{-2}\bar{X}(\hat{n}) \,.
\eeqry
\end{subequations}
%
%which by the virtue of being scalars are independent of coordinate definitions. %The real part of this complex spin-0 scalar field corresponds to the E-mode while the imaginary part to the B-modes \cite{Kamionkowski1997}. 
The $\cal E/B$ fields are defined locally at point $\hat n$ in terms of the operators $\eth$ and $\bar \eth$.

It is possible to decompose the complex field $_{\pm 2}X$ into spin spherical harmonic functions: ${}_{\pm 2}X(\hat{n}) = \sum_{\ell m} {}_{\pm 2} X_{\ell m} {}_{\pm 2}Y_{\ell m}(\hat{n})$. On applying the spin raising and lowering operators on the spin spherical harmonic functions leads to the following identities \cite{goldberg67}:
%
\begin{subequations}\label{eq:spinopylm} 
\beqry
\eth _s Y_{lm}(\hat{n}) &=& \sqrt{(\ell-s)(\ell+s+1)} _{s+1} Y_{lm}(\hat{n}) \,, \\
\bar{\eth} _s Y_{lm}(\hat{n}) &=& -\sqrt{(\ell+s)(\ell-s+1)} _{s-1} Y_{lm}(\hat{n}) \,, 
\eeqry
\end{subequations}
%
where $_s Y_{lm}(\hat{n}) $ denote the spin-s spherical harmonics.

From the definition of $\mathcal{E/B}$, the spin spherical harmonic decomposition of ${}_{\pm2}X$, and the identities given in \eq{eq:spinopylm}, it follows that the scalar fields $\mathcal{E}/\mathcal{B}$ are given by the equations:
%
\beq \label{eq:pseudo}
\mathcal{E}(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n})\qquad;\qquad
\mathcal{B}(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) \,,
\eeq
%
where the harmonic coefficients $a^{E}_{\ell m}$ and  $a^{B}_{\ell m}$ relate to the harmonic coefficients of the spin-2 polarization field via the following equations:
%
\beq\label{eq:x2eb}
a^{E}_{\ell m} = -\frac{1}{2} \Big[ {}_{+2}\tilde{X}_{\ell m} + {}_{-2}\tilde{X}_{\ell m} \Big]\qquad;\qquad a^{B}_{\ell m} = -\frac{1}{2i} \Big[ {}_{+2}\tilde{X}_{\ell m} - {}_{-2}\tilde{X}_{\ell m} \Big] \,.
\eeq
%
In the remainder of this article, we will work with the scalar $E$ and pseudo scalar $B$ fields, defined by: 
%
\beq \label{eq:realeb}
E(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} Y_{\ell m} (\hat{n})\qquad;\qquad B(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} Y_{\ell m} (\hat{n}) \,.
\eeq
%
These $E/B$ fields are merely red-filtered versions of $\mathcal{E}/\mathcal{B}$ (their spherical harmonic coefficients of expansion are related by the factor $[{(\ell+2)!}/{(\ell-2)!}]^{1/2}$) and are not local functions of Stokes $Q/U$. %We make this choice since the CMB spectra are more closely related to the fields E \& B.

\subsection{Matrix notation} \label{sec:mat_pol_intro}
Our derivation of real space operators is more transparent in a matrix-vector notation.\footnote{While we work with the matrix and vector sizes given in terms of some pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}
We introduce a matrix that encodes spin spherical harmonic basis vectors:
%
\beq
{}_{|s|}\mathcal{Y}= \bmat _{+s}Y & 0 \\ 0 & _{-s}Y \emat _{2 \rm N_{\rm pix} \times 2 \rm N_{\rm alms}} \,.
\eeq
%
%where $s$ denotes the spin of the basis functions.
We will be working with cases $s \in [0,2]$. Each column maps to a specific harmonic basis function (i.e. indexed by $\ell m$) and each row maps to a pixel on the sphere. This matrix is not square in general: the number of rows is determined by the pixelization and the number of columns is set by the number of basis functions (e.g. determined by the band limit).

@@

We define transformations between different representations of the polarization field (i.e. from $Q,U$ to $_{\pm2} X$ and back):
%
\beqrys
\bar T &=& \qutox_{2 \rm N_{\rm pix} \times 2 \rm N_{\rm pix}} ;\qquad \bar T^{-1} = \frac{1}{2} \bar T^{\dagger} \,, \\
\tilde T &=& -\qutox_{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms}};\qquad \tilde T^{-1} = \frac{1}{2} \tilde T^{\dagger} \,,
\eeqrys
%
The sign conventions we have chosen match HEALPix.
Using the data vectors and the matrix operators defined above we can now express, in compact notation, the forward and inverse relations between different representations of the polarization data vectors via the following equations:
%
\begin{subequations} \label{eq:pol_data_relns}
  %Kevin's version
  \beqry
  \bar{X} &= \bar T  \bar{P} \qquad; &\qquad \bar{P} = \frac{1}{2} \bar T^{\dagger}  \bar{X} \,, \\
  \tilde{X} &= \tilde T \tilde{S} \qquad; &\qquad \tilde{S} = \frac{1}{2}\tilde T^{\dagger} \tilde{X} \,.
  \eeqry
  Meanwhile the spherical harmonic transforms are written as:
  \beqry
  \bar X &=  {{}_2\mathcal{Y}}  \tilde X; &\qquad \tilde X ={{}_2\mathcal{Y}}^{\dagger}  \bar X  ; \\
  \bar S &=  {{}_0\mathcal{Y}} \tilde S; &\qquad  \tilde S =  {{}_0\mathcal{Y}}^{\dagger} \bar S \,.
  \eeqry
% Aditya's version
%\beqry
%\bar{X} &=& \bar T * \bar{P} ~~;~~\bar{P} = \frac{1}{2} \bar T^{\dagger} * \bar{X} \,, \\
%\bar X &=&  {{}_2\mathcal{Y}} * \tilde X  ~~;~~ \tilde X ={{}_2\mathcal{Y}}^{\dagger} * \bar X  \,, \\
%\tilde{X} &=& \tilde T * \tilde{S} ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,.\\ 
%\bar S &=&  {{}_0\mathcal{Y}} * \tilde S ~~;~~  \tilde S =  {{}_0\mathcal{Y}}^{\dagger} * \bar S \,.
%\tilde X &=&  {{}_2\mathcal{Y}}^{\dagger} * \bar X ~~;~~ \tilde{X} = \tilde T * \tilde{S} \,, \\
%\bar{S} &=& {{}_0\mathcal{Y}}*\tilde S ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,.\\
%\bar{X} &=& \bar T * \bar{P} ~~;~~ \tilde{X} = \tilde T * \tilde{S} \,, \\
%\bar{P} &=& \frac{1}{2} \bar T^{\dagger} * \bar{X} ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,. \\
%\eeqry
\end{subequations}
%
Finally we introduce the operators that project harmonic space data vector to the $E$ or $B$ subspace:
%
\begin{subequations} \label{eq:har_eb_op}
\beqry
\tilde O_E &=& \bmat \mathbb{1} & \mathbb{0} \\ \mathbb{0} & \mathbb{0} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} }   ~~;~~ \tilde S_E = \tilde O_E  \tilde S ,\\
\tilde O_B &=& \bmat \mathbb{0} & \mathbb{0} \\ \mathbb{0} & \mathbb{1} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} } ~~; ~~ \tilde S_B = \tilde O_B  \tilde S .
\eeqry
\end{subequations}
%
Note that these harmonic space matrices are idempotent: ($\tilde O_E  \tilde O_E = \tilde O_E;  \tilde O_B  \tilde O_B= \tilde O_B$), orthogonal: ($\tilde O_E  \tilde O_B = \mathbb{0}$), and sum to the identity matrix: ($\tilde O_E + \tilde O_B = \mathbb{1}$).
%
%\begin{subequations} \label{eq:har_op_prop}
%\beqry
%\tilde O_E  \tilde O_E&=& \tilde O_E ~~;~~  \tilde O_B  \tilde O_B= \tilde O_B \,,\\
% \tilde O_E  \tilde O_B&=& \mathbb{0} \,, \label{eq:op_eb_ortho}\\ 
% \tilde O_E + \tilde O_B&=& \mathbb{1} \,.
%\eeqry
%\end{subequations}
%
The above relations for these harmonic space operators are exactly valid.  In the following sections we derive the real space analogues ($O_E,O_B$) of these harmonic space operators.

>>>>

\section{Polarization primer} \label{sec:pol-primer}
%\subsection{Constructing $E/B$ fields without harmonic space}
\subsection{A heuristic argument for real space construction of E/B fields on the sphere} \label{sec:qu2eb_heuristic}
%\comment{Need a better title for the section that gives the heuristic argument}
%\textit{Alternative for bringing the heuristic argument earlier.}

 CMB polarization is measured in terms of Stokes parameters, time averages of the linear polarization of the electric field along cartesian axes perpendicular to the line of sight.\footnote{Throughout we use the conventions of HEALPix \cite{healpix_primer}, measuring the polarization angle East of South.} Thus Stokes $Q$ and $U$ depend on the choice of the local coordinate system and a rotation by an angle $\psi$ around the line of sight transforms them as:
%
\beq \label{eq:qu-rot}
\fqu' = \begin{bmatrix} \cos{2 \psi} &  \sin{2 \psi} \\ -\sin{2\psi} & \cos{2 \psi} \end{bmatrix} \fqu \,.
\eeq
Equivalently the object $_{\pm 2}\bar{X}(\hat{n}) = Q(\hat{n}) \pm i U (\hat{n})$ transforms as ${}_{\pm 2}f' = e^{\mp 2i\psi} {}_{s}f$ and hence forms a spin-2 field \cite{Zaldarriaga1997}.

@@

We consider the contribution to a scalar field at $\hat n_e$ from the polarization field at $\hat n_q$. \fig{fig:euler_angles} shows that the transformation of the local coordinate system between the two positions can be described by a \revisit{counter-clockwise rotation around the local $\hat n_q$ (unit outward pointing radial vector at $(\theta_q,\phi_q)$) by angle $\alpha$, parallel transport by angle $\beta$ along the shorter geodesic and a counter-clockwise rotation around $\hat n_q$ by $-\gamma$.  This corresponds to a rotation by Euler angles $(\alpha,\beta,-\gamma)$ in the $z-y_1-z_2$ convention.\footnote{The Euler angles in the more standard $z-y-z$ convention are related to those in the $z-y_1-z_2$ convection by the following rule: $(\alpha,\beta,\gamma)_{z-y-z} =(\gamma,\beta,\alpha)_{z-y_1-z_2}$ \cite{varshalovich}.}}
%
\begin{figure}%[!hbt]
\centering
\includegraphics[width=0.5\columnwidth]{euler.pdf}
\caption{This figure depicts the Euler angles in the z-y1-z2 convention. The cartesian coordinates shown in \revisit{dark solid green} are those that lie in the tangent plane at location $\hat{n}_q = (\theta_q, \phi_q)$ while those shown in \revisit{dark solid red} are the ones that lie in the tangent plane at location $\hat{n}_e = (\theta_e, \phi_e)$. The blue coordinates at different locations are representative of the parallel transport along the geodesic connection the two locations $\hat{n}_q$ \& $\hat{n}_e$ on the sphere.}
\label{fig:euler_angles}
\end{figure}
%

Rotating the cartesian coordinates in the tangent plane at location $\hat{n}_q$ by an angle $\psi$ about the local $\hat{z}_q$ axis, the Stokes parameters in the new coordinate system relate to those in the original coordinate system as:
$\mathcal{R}_{\hat{z}_q}(\psi)[{}_{+2}X(\hat{n}_q)] =  {}_{+2}X(\hat{n}_q) e^{-i2\psi} $.
This same rotation by $\psi$ alters the Euler angle $\alpha_{qe}$ (that aligns the $x$-axis at $\hat{n}_q$ along the geodesic to the location $\hat{n}_e$): $\mathcal{R}_{\hat{z}_q}(\psi)[\alpha_{qe}] = \alpha_{qe} - \psi$.  Therefore one can see that $\mathcal{R}_{\hat{z}_q}(\psi)[e^{-i2\alpha_{qe}}] =  e^{-i2\alpha_{qe}} e^{i2\psi}$.

Given these transformation properties, the combination ${}_{+2}X(\hat{n}_q)e^{-i2\alpha_{qe}}$ is invariant under rotations and must be spin-0 by definition:
\beq
\mathcal{R}_{\hat{z}_q}(\psi)[{}_{+2}X(\hat{n}_q)e^{-i2\alpha'_{qe}}] = {}_{+2}X(\hat{n}_q) e^{-i2\alpha_{qe}} \,. \label{eq:invariant}
\eeq
Further note that $Q~ \& \cos2 \alpha$ have even parity since they do not change sign when $\hat{x} \rightarrow -\hat{x}$ (or $\hat{y} \rightarrow -\hat{y}$  ) while $U~ \& \sin 2 \alpha$ change sign under this transformation and hence have odd parity. The real part of the function  ${}_{+2}X(\hat{n}_q)e^{-i2\alpha_{qe}}$ is composed of $Q\cos{2\alpha} ~ \& ~ U\sin{2 \alpha}$ which are product of functions having the same parity and hence the real part must have even parity. Similarly the imaginary part of the function is composed of $Q\sin{2 \alpha}~\&~ U\cos{2\alpha}$ which are product of functions having opposite parity and hence the imaginary part must have an odd parity. Therefore we can make the association that contributions to  $(E+iB)(\hat n_e)$ must be proportional to $ {}_{+2}X(\hat{n}_q) e^{-i2\alpha_{qe}}$.

The same rotation $\mathcal{R}_{\hat{z}_q}(\psi)$ leaves the Euler angle $|\beta_{qe}|$ unaltered (it measures the angular distance between the points).  Thus we can conclude that the contribution to $(E+iB)(\hat n_e)$ from the position $n_q$ must have the form:
\beq
{}_{+2}X(\hat{n}_q) f(\beta_{qe})  e^{-i2\alpha_{qe}}
\eeq
for some real function $f$.  Note that when the two locations coincide ($\beta_{qe}=0$) then  $\alpha_{qe}=0,2\pi,4\pi,\dots$, implying $E + iB \propto Q+iU$.  This is a contradiction because $Q+iU$ does not transform as a spin-0 field under local rotations, and so we must have $f(\beta_{qe} = 0 ) = 0$.   Hence the $E/B$ fields are necessarily non-local.  A similar contradiction arises when the two locations are diametrically opposite, $\beta_{qe} = \pi$, and so $f(\beta_{qe} = \pi ) = 0$.  Any such function $f$ will let us construct $E/B$-like scalar fields.  Below we derive the particular one that gives rise to our familiar $E/B$ modes.

@@

The standard construction of $E/B$ fields depend on the spin-raising and -lowering operators, and is usually carried out in harmonic space.
%
The spin-raising operator ($\eth$), applied to a field of spin-s $_{s}g$, results in a field with spin-$(s+1)$: $(\eth _{s}g)' = e^{-i(s+1)\psi}(\eth _{s}g)$  \cite{goldberg67}.  The complementary spin-lowering operator $(\bar{\eth})$  similarly results in a field with spin-$(s-1)$: $(\bar{\eth} _{s}g)' = e^{-i(s-1)\psi}(\bar{\eth} _{s}g)$.  The complex spin-0 scalar now arise from  the spin-2 fields ${_{\pm 2}X}$ as follows,
%
\begin{subequations}\label{eq:ebdef}
\beqry
\mathcal{E}(\hat{n}) + i \mathcal{B}(\hat{n}) &=& -\bar{\eth}^2 _{+ 2}\bar{X}(\hat{n}) \,,\label{eq:ebdef_lower}\\
\mathcal{E}(\hat{n}) - i \mathcal{B}(\hat{n}) &=& -{\eth}^2 _{-2}\bar{X}(\hat{n}) \,.
\eeqry
\end{subequations}
%
%which by the virtue of being scalars are independent of coordinate definitions. %The real part of this complex spin-0 scalar field corresponds to the E-mode while the imaginary part to the B-modes \cite{Kamionkowski1997}. 
The $\cal E/B$ fields are defined locally at point $\hat n$ in terms of the operators $\eth$ and $\bar \eth$.

It is possible to decompose the complex field $_{\pm 2}\bar{X}$ into spin spherical harmonic functions: ${}_{\pm 2}\bar{X}(\hat{n}) = \sum_{\ell m} {}_{\pm 2} X_{\ell m} {}_{\pm 2}Y_{\ell m}(\hat{n})$. Applying the spin raising and lowering operators on the spin spherical harmonic functions leads to the following identities \cite{goldberg67}:
%
\begin{subequations}\label{eq:spinopylm} 
\beqry
\eth _s Y_{lm}(\hat{n}) &=& \sqrt{(\ell-s)(\ell+s+1)} _{s+1} Y_{lm}(\hat{n}) \,, \\
\bar{\eth} _s Y_{lm}(\hat{n}) &=& -\sqrt{(\ell+s)(\ell-s+1)} _{s-1} Y_{lm}(\hat{n}) \,, 
\eeqry
\end{subequations}
%
where $_s Y_{lm}(\hat{n}) $ denote the spin-s spherical harmonics.

From the definition of $\mathcal{E/B}$, the spin spherical harmonic decomposition of ${}_{\pm2}\bar{X}$, and the identities given in \eq{eq:spinopylm}, it follows that the scalar fields $\mathcal{E}/\mathcal{B}$ are given by the equations:
%
\beq \label{eq:pseudo}
\mathcal{E}(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n})\qquad;\qquad
\mathcal{B}(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) \,,
\eeq
%
where the harmonic coefficients $a^{E}_{\ell m}$ and  $a^{B}_{\ell m}$ relate to the harmonic coefficients of the spin-2 polarization field via the following equations:
%
\beq\label{eq:x2eb}
a^{E}_{\ell m} = -\frac{1}{2} \Big[ {}_{+2}\tilde{X}_{\ell m} + {}_{-2}\tilde{X}_{\ell m} \Big]\qquad;\qquad a^{B}_{\ell m} = -\frac{1}{2i} \Big[ {}_{+2}\tilde{X}_{\ell m} - {}_{-2}\tilde{X}_{\ell m} \Big] \,.
\eeq
%
In the remainder of this article, we will work with the scalar $E$ and pseudo scalar $B$ fields, defined by: 
%
\beq \label{eq:realeb}
E(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} Y_{\ell m} (\hat{n})\qquad;\qquad B(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} Y_{\ell m} (\hat{n}) \,.
\eeq
%
These $E/B$ fields are merely \revisit{red-filtered} \comment{Is this unabbiguously understood as down weighing the high multipoles ? } versions of $\mathcal{E}/\mathcal{B}$ (their spherical harmonic coefficients of expansion are related by the factor $[{(\ell+2)!}/{(\ell-2)!}]^{1/2}$) and are not local functions of Stokes $Q/U$. %We make this choice since the CMB spectra are more closely related to the fields E \& B.

\subsection{Matrix notation} \label{sec:mat_pol_intro}
Our derivation of real space operators is more transparent in a matrix-vector notation.\footnote{While we work with the matrix and vector sizes given in terms of some pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}
We introduce a matrix that encodes spin spherical harmonic basis vectors:
%
\beq
{}_{|s|}\mathcal{Y}= \bmat _{+s}Y & 0 \\ 0 & _{-s}Y \emat _{2 \rm N_{\rm pix} \times 2 \rm N_{\rm alms}} ;\qquad  {}_{|s|}\mathcal{Y}^{\dagger}= \Delta \Omega \bmat _{+s}Y^{\dagger} & 0 \\ 0 & _{-s}Y^{\dagger} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm pix}} \,,
\eeq
%
\revisit{where $s$ denotes the spin of the basis functions and our definition of ${}_{|s|}\mathcal{Y}^{\dagger}$ differs from the conventional dagger operation by the factor $\Delta \Omega$ which is introduced to ensures the ortho-\textit{normality} of the basis functions: ${}_{|s|}\mathcal{Y}^{\dagger} {}_{|s|}\mathcal{Y} = I_{2 \rm N_{\rm alms} \times2 \rm N_{\rm alms}}$ on the discretized sphere.}

We will be working with cases $s \in [0,2]$. Each column of ${}_{|s|}\mathcal{Y}$ maps to a specific harmonic basis function (i.e. indexed by $\ell m$) and each row maps to a pixel on the sphere. This matrix is not square in general: the number of rows is determined by the pixelization and the number of columns is set by the number of basis functions (e.g. determined by the band limit).

@@

We define transformations between different representations of the polarization field (i.e. from $Q,U$ to $_{\pm2}\bar{X}$ and back):
%
\beqrys
\bar T &=& \qutox_{2 \rm N_{\rm pix} \times 2 \rm N_{\rm pix}} ;\qquad \bar T^{-1} = \frac{1}{2} \bar T^{\dagger} \,, \\
\tilde T &=& -\qutox_{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms}};\qquad \tilde T^{-1} = \frac{1}{2} \tilde T^{\dagger} \,,
\eeqrys
%
The sign conventions we have chosen match HEALPix.
Using the data vectors and the matrix operators defined above we can now express, in compact notation, the forward and inverse relations between different representations of the polarization data vectors via the following equations:
%
\begin{subequations} \label{eq:pol_data_relns}
  %Kevin's version
  \beqry
  \bar{X} &= \bar T  \bar{P} \qquad; &\qquad \bar{P} = \frac{1}{2} \bar T^{\dagger}  \bar{X} \,, \\
  \tilde{X} &= \tilde T \tilde{S} \qquad; &\qquad \tilde{S} = \frac{1}{2}\tilde T^{\dagger} \tilde{X} \,.
  \eeqry
  Meanwhile the spherical harmonic transforms are written as:
  \beqry
  \bar X &=  {{}_2\mathcal{Y}}  \tilde X; &\qquad \tilde X ={{}_2\mathcal{Y}}^{\dagger}  \bar X  ; \\
  \bar S &=  {{}_0\mathcal{Y}} \tilde S; &\qquad  \tilde S =  {{}_0\mathcal{Y}}^{\dagger} \bar S \,.
  \eeqry
% Aditya's version
%\beqry
%\bar{X} &=& \bar T * \bar{P} ~~;~~\bar{P} = \frac{1}{2} \bar T^{\dagger} * \bar{X} \,, \\
%\bar X &=&  {{}_2\mathcal{Y}} * \tilde X  ~~;~~ \tilde X ={{}_2\mathcal{Y}}^{\dagger} * \bar X  \,, \\
%\tilde{X} &=& \tilde T * \tilde{S} ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,.\\ 
%\bar S &=&  {{}_0\mathcal{Y}} * \tilde S ~~;~~  \tilde S =  {{}_0\mathcal{Y}}^{\dagger} * \bar S \,.
%\tilde X &=&  {{}_2\mathcal{Y}}^{\dagger} * \bar X ~~;~~ \tilde{X} = \tilde T * \tilde{S} \,, \\
%\bar{S} &=& {{}_0\mathcal{Y}}*\tilde S ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,.\\
%\bar{X} &=& \bar T * \bar{P} ~~;~~ \tilde{X} = \tilde T * \tilde{S} \,, \\
%\bar{P} &=& \frac{1}{2} \bar T^{\dagger} * \bar{X} ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,. \\
%\eeqry
\end{subequations}
%
Finally we introduce the operators that project harmonic space data vector to the $E$ or $B$ subspace:
%
\begin{subequations} \label{eq:har_eb_op}
\beqry
\tilde O_E &=& \bmat \mathbb{1} & \mathbb{0} \\ \mathbb{0} & \mathbb{0} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} }   ~~;~~ \tilde S_E = \tilde O_E  \tilde S ,\\
\tilde O_B &=& \bmat \mathbb{0} & \mathbb{0} \\ \mathbb{0} & \mathbb{1} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} } ~~; ~~ \tilde S_B = \tilde O_B  \tilde S .
\eeqry
\end{subequations}
%
Note that these harmonic space matrices are idempotent: ($\tilde O_E  \tilde O_E = \tilde O_E;  \tilde O_B  \tilde O_B= \tilde O_B$), orthogonal: ($\tilde O_E  \tilde O_B = \mathbb{0}$), and sum to the identity matrix: ($\tilde O_E + \tilde O_B = \mathbb{1}$).
%
%\begin{subequations} \label{eq:har_op_prop}
%\beqry
%\tilde O_E  \tilde O_E&=& \tilde O_E ~~;~~  \tilde O_B  \tilde O_B= \tilde O_B \,,\\
% \tilde O_E  \tilde O_B&=& \mathbb{0} \,, \label{eq:op_eb_ortho}\\ 
% \tilde O_E + \tilde O_B&=& \mathbb{1} \,.
%\eeqry
%\end{subequations}
%
The above relations for these harmonic space operators are exactly valid.  In the following sections we derive the real space analogues ($O_E,O_B$) of these harmonic space operators.
